A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution

### 2013-03-12 ISMB/results_ISMB.tex
intermediate commit (results)

\subsection{Design}
 To benchmark our method, we first started by sampling about $100$ sequences per structure. After correcting the free nucleotides with
 \RNAinverse we computed their MFE with \rnafold from the \textit{Vienna Package 2.0} by~\cite{Hofacker:1994}.

A first criterion is the proportion of sequences accepting the target
secondary structures as their MFE. A second criterion is the number of structures
for which at least one sequence with the desired MFE was produced.
Figure~\ref{fig:mfe_struct_solved_noinverse} shows the results obtained by \ourprog for these criteria. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \ourprog sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
summarized by Figure~\ref{fig:mfe_struct_solved}. We observed
that length is, in general, not a good predictor for the hardness of designing a structure. 
Instead, a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
 The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last column of Figure~\ref{fig:mfe_struct_solved}.
Even with a \GCContent of $50\%$ and more, when at least
one solution was found for almost all structures, most of the samples 
do not fold into the target. A few structures remained unsolved under
all \GCContent.

\begin{figure*}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solved_vsrnainverse}
	\caption{Quality of \ourprog results. The first row shows the percentage
	of sampled sequences folding into the target when using only \ourprog. The second shows 
  after processing previous results with \RNAinverse.
	}
	\label{fig:mfe_struct_solved_noinverse}	
\end{figure*}

@@

To evaluate the global quality of \ourprog sequences, we show
in Figure~\ref{fig:ss_sens} the ratio of well predicted base pairs in the
MFE structure of our sampled sequences. We can observe that, in all cases,
the hardest sequence to design are those with an extremely low \GCContent. As anticipated, those are the sequences with the weakest bonds.
We notice that the most accurate sequences yield a \GCContent
of $70\pm 10\%$.

As discussed in Sec.~\ref{sec:implementation}, we notice a highly decreased
computational time needed to generate the sequences with $15\%$ free 
nucleotides in the loops. We remark that those structures also yield 
a much lower structural sensitivity.

@@

\subsection{Samples properties}

In this section, we further analyze the generated sequences that fold into the 
target structure.

Using the same dataset of $50$ structures, we generated $100$ samples
per structure with classical softwares who can not control the \GC content.%with \RNAinverse. They yield for most parameters
%a high entropy.
%Table.~\ref{tab:nb_rnainv} contains the number of samples with a given \GCContent generated
with  \RNAinverse, \INFORNA, \NUPACK and \frankenstein.  We present in Fig.~\ref{fig:gcdist} those distributions. They have a highly concentrated distribution of \GC content showing one of theirs major deficiencies, with no sequence below $30\%$\GC. Therefore the need to be able to explicitly
 control the \GC content.

@@

A desirable feature in sequence design, is to produce samples with a high
diversity and stable secondary structure. 
Two useful measures are the sequence identity Boltzmann frequency.
%The positional entropy is defined as:
%\begin{align}
%  \sum_{\substack{i\\b \in (\Ab,\Cb,\Gb,\Ub)}} -p_{i\to b}\cdot \log_4(p_{i\to b}) &&& \text{(Positional entropy)}
%  \label{eq:entropy}
%\end{align}
%where $p_{i\to b}$ is the proportion of nucleotides $b$ at position $i$.
%Intuitively, this measure captures the variability of sequences generated by a given method.
The sequence identity is defined over a set $\mathcal{S}$ of sequences of equal length as follows:
\begin{align}
  \sum_{s^1,s^2 \in\mathcal{S}\times\mathcal{S}}\left(\frac{1}{|s^1|}{\sum_{\substack{i\\s^1_i\equiv s^2_i}}1}\right) &&& \text{Seq. identity}
\end{align}
where $s_i$ is the nucleotide at position $i$ in sequence $s$.
Intuitively, this measure captures the variability of sequences generated by a given method.
The Boltzmann frequency is defined, for a structure $\Struct$ and a sequence $s$ as:
\begin{align}
  e^{\frac{-\ES(s,\Struct)}{RT}}/\mathcal{Z}^s  &&&\text{Frequency}
\end{align}
where $\mathcal{Z}^s$ is the partition function of sequence $s$. This measure tells us how dominant is structure $\Struct$ in the Boltzmann ensemble of sequence $s$. A high value implies a 
stable structure. To compute the frequency, we used \rnafold from the \texttt{Vienna Package 2.0} developed by~\cite{Hofacker:1994}.

>>>>

 \highlight{
\subsection{Design}
We ran our method as follows. First, we sampled approximately $100$ sequences per structure. Then, we use these sequences 
as seed in \RNAinverse. Finally, we computed the MFE with the \rnafold program from the \textit{Vienna Package 2.0} \citep{Hofacker:1994}.

\subsection{Success rate}
We started by estimating the success rate of our methodology and computed the percentage of sequences with a MFE structure identical to the target secondary structure. Figure~\ref{fig:mfe_struct_solved_noinverse} shows our results. We clearly see that before the post-processing step (i.e. \RNAinverse) the sequences sampled by \ourprog have a low success rate (first row). As mentioned earlier, this could be explained by the fact that no selection criterion has been at this stage applied to unpaired nucleotides. Remarkably, after the local search optimization (with \RNAinverse) of nucleotides in unpaired regions (second row), we observe a dramatic improvement of our success rate. As expected, we observed that length is, in general, not a good predictor for the hardness of designing a structure. Instead, a high number of free nucleotides in the structure seems to be a  good measure of the hardness of its design. Similarly, these data also show that designing sequences with low \GCContent is challenging for all types of targets.
}

\begin{figure*}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solved_vsrnainverse}
	\caption{Success rate \ourprog before after after \RNAinverse post-processing. The first row shows the percentage of sampled sequences folding into the target when using only \ourprog. The second shows after processing previous results with \RNAinverse.}
	\label{fig:mfe_struct_solved_noinverse}	
\end{figure*}

@@

\highlight{ 
We investigated further the quality of the sequences generated by \ourprog. In particular, we estimated the capacity of our methods to generate ``good'' sequences with desired folding capabilities regardless of the property  to fold exactly into the target structure. In Figure~\ref{fig:ss_sens}, we show the ratio of well predicted base pairs in the MFE structure of our sampled sequences. As above, we can observe that, in all cases, the sequences that are the hardest to design are those with an extremely low \GCContent. Indeed, the energetic contribution of the base pairs to the stability of the structure is weaker. Interestingly, we also notice that the most accurate sequences yield a \GCContent of $70\pm 10\%$. Overall, we observe that all our samples have good folding properties, and that there is a correlation between the ``precision'' of the samples and the hardness of the design.
}

As discussed in Sec.~\ref{sec:implementation}, we noticed a highly decreased computational time needed to generate the sequences with $15\%$ free 
nucleotides in the loops. We remark that those structures also yield a much lower structural sensitivity.

@@

\subsection{Properties of designed sequences}

In this section, we further analyze the generated sequences with a MFE structure that folds into the target structure.

\highlight{
Using the same dataset of $50$ structures, we generated $100$ samples per structure with classical softwares who do not control the \GCContent. Namely,  \RNAinverse, \INFORNA, \NUPACK and \frankenstein.  We show the distribution of the \GCContent of the sequences produced with these softwares in Fig.~\ref{fig:gcdist} those distributions. As anticipated, we observe a clear bias toward high {\GCContent}s and a complete absence of sequence with less than $30\%$ of \GC. This striking results motivates a need for methods that enable to explicitly control the \GCContent and more precisely that enable to design sequences with low \GCContent (i.e. 30\% or less).
}

@@

\highlight{
A desirable feature in sequence design, is to produce samples with a high sequence diversity and stable secondary structure. 
Therefore, in the following we will use two useful measures which are the sequence identity of the samples, and the Boltzmann probability
of the target structure in the low energy ensemble.
}
%The positional entropy is defined as:
%\begin{align}
%  \sum_{\substack{i\\b \in (\Ab,\Cb,\Gb,\Ub)}} -p_{i\to b}\cdot \log_4(p_{i\to b}) &&& \text{(Positional entropy)}
%  \label{eq:entropy}
%\end{align}
%where $p_{i\to b}$ is the proportion of nucleotides $b$ at position $i$.
%Intuitively, this measure captures the variability of sequences generated by a given method.
\highlight{
The sequence identity is defined over a set $\mathcal{S}$ of aligned sequences (in our case, all sequences have the same length and can be trivially aligned) as :
\begin{align}
  \sum_{s^1,s^2 \in\mathcal{S}\times\mathcal{S}}\left(\frac{1}{|s^1|}{\sum_{\substack{i\\s^1_i\equiv s^2_i}}1}\right) &&& \text{Seq. identity}
\end{align}
where $s_i$ is the nucleotide at position $i$ in sequence $s$. Intuitively, this measure captures the diversity of sequences generated by a given method. Next, the Boltzmann frequency is defined, for a structure $\Struct$ and a sequence $s$ as:
\begin{align}
  e^{\frac{-\ES(s,\Struct)}{RT}}/\mathcal{Z}^s  &&&\text{Frequency}
\end{align}
where $\mathcal{Z}^s$ is the partition function of sequence $s$. This measure tells us how dominant is a structure $\Struct$ in the Boltzmann ensemble of structure over a sequence $s$. A high value implies a stable structure. We compute this frequency with \rnafold from the \texttt{Vienna Package 2.0} \citep{Hofacker:1994}.
}
